---
title: "Dynamic vs sprint bioenergetic model on a 200 m sprint"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Dynamic vs sprint bioenergetic model on a 200 m sprint}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4
)
```

```{r setup}
library(runrgetics)
library(dplyr)
library(ggplot2)
```

## Two models, one sprint

`runrgetics` decomposes a sprint's metabolic power into aerobic,
anaerobic-alactic and anaerobic-lactic contributions with the **sprint
bioenergetic model** (`analyze_sprint_bioenergetics()`): a set of empirical
power-time curves (a log-normal alactic bump, a bi-exponential lactic term and an
exponential aerobic rise) fitted to the di Prampero metabolic-power demand.

This vignette adds a second, mechanistically different lens on the same data: a
**dynamic Mader-style metabolic model**, ported from the
[MuscleEdot](https://github.com/) package into `runrgetics`
(`simulate_dynamic_bioenergetics()`). Instead of fitting empirical curves, it
*integrates a 5-state system of ordinary differential equations* (high-energy
phosphates, muscle oxygen uptake, muscle and blood lactate, cumulative ATP demand)
driven by the measured power demand, with oxidative phosphorylation, glycolysis
and lactate oxidation controlled by the cytosolic phosphorylation state (free
ADP) and muscle pH. It then reads off the instantaneous aerobic, glycolytic and
alactic power the same way the ECSS energy-partition figure does.

The two models are driven by **exactly the same observed signal** — the di
Prampero sprint cost of running times velocity — so their partitions are directly
comparable. VO2max is fixed and VLamax is tuned to the demand. The question is
whether a dynamic, ADP/pH-regulated model recovers a partition comparable to the
empirical sprint model on a real 200 m repetition.

## The target sprint

We use the ten-200 m paired session and its gpexe reference signal, and take the
first detected repetition (`sprint_id = 1`), exactly as the sprint bioenergetic
model does. Body mass for this athlete is 67 kg; the session maximal aerobic
power (MAP) is 27 W/kg.

```{r detect}
gpexe <- subset(ten_200_sprints_paired, source == "gpexe")
sprints <- detect_sprints(gpexe)
sprints[1, c("sprint_id", "duration", "distance", "peak_speed")]
```

The sprint bioenergetic model's decomposition of this sprint:

```{r sprint-model}
plot_sprint_bioenergetics(gpexe, sprint_id = 1, maximal_aerobic_power = 27)
```

## Parameters of the dynamic model

The dynamic model works in the same units as the rest of `runrgetics` (power in
W/kg, time in s), and takes systemic VO2max (mL/kg/min) and VLamax (mmol/kg/s),
converting them to muscle-specific values internally.

* **VO2max is fixed.** The athlete's measured VO2max is 75 mL/kg/min. The dynamic
  model's VO2max corresponds to the measured value over-estimated by ~5 % (the
  model's oxidative machinery never quite reaches the whole-body plateau), i.e.
  `75 * 1.05 = 78.75` mL/kg/min.
* **VLamax is the free parameter**, tuned to minimise the RMSE between the
  dynamic model's total metabolic power and the di Prampero demand.
* **SC, SA and M_active** are held at the MuscleEdot defaults (25 mmol/kg,
  7 mmol/kg and 0.25).

## Tuning VLamax to the di Prampero demand

`tune_dynamic_vlamax()` does a 1-D optimisation of VLamax against the sprint's
metabolic-power demand. `compare_bioenergetic_models()` wraps the whole pipeline
(detect the sprint, build the demand, fit the sprint model, tune and run the
dynamic model) and, when `vlamax` is left `NULL`, tunes it for us.

```{r compare}
vo2max <- 75 * 1.05 # ~78.75 mL/kg/min, fixed
cmp <- compare_bioenergetic_models(
  gpexe, sprint_id = 1, maximal_aerobic_power = 27, vo2max = vo2max
)

tuned_vlamax <- attr(cmp, "vlamax")
tuned_rmse <- attr(cmp, "rmse")
c(vlamax = tuned_vlamax, rmse = tuned_rmse)
```

The tuned VLamax lands in the low-to-mid physiological range, and the dynamic
model's total metabolic power tracks the di Prampero demand to within an RMSE of
about `r round(tuned_rmse, 2)` W/kg over the whole sprint.

## Side-by-side partition

`compare_bioenergetic_models()` returns a tidy table of each pathway's energy
(J/kg) and percentage share for both models.

```{r table}
cmp
```

```{r table-wide}
# a compact per-pathway comparison of the two models' percentage shares
sprint_pct <- with(cmp[cmp$model == "sprint", ], stats::setNames(pct, pathway))
dynamic_pct <- with(cmp[cmp$model == "dynamic", ], stats::setNames(pct, pathway))
data.frame(
  pathway = levels(cmp$pathway),
  sprint_pct = round(sprint_pct[levels(cmp$pathway)], 1),
  dynamic_pct = round(dynamic_pct[levels(cmp$pathway)], 1),
  row.names = NULL
)
```

## Overlaying the power-time curves

```{r plot-overlay}
plot_bioenergetic_comparison(
  gpexe, sprint_id = 1, maximal_aerobic_power = 27, vo2max = vo2max
)
```

## Where the models agree and diverge

Two things stand out on this 200 m sprint:

* **Total energy and the aerobic share agree well.** Both models reconstruct a
  similar total metabolic energy for the sprint, and both attribute a comparable
  fraction (~40 %) to aerobic metabolism — a reassuring, model-independent result
  given how differently the aerobic term is built (an empirical exponential rise
  vs. an ADP-driven Hill function with first-order VO2 kinetics).

* **They split the anaerobic budget differently.** The empirical sprint model is
  *lactic-heavy* (a large bi-exponential lactic term, a small alactic bump),
  whereas the dynamic model is *alactic-heavy*: its phosphocreatine pool covers
  most of the early, high-power transient before glycolysis and oxidative
  phosphorylation take over. This is the expected signature of the two
  formulations. The sprint model places the alactic contribution in a narrow
  log-normal bump around the launch, while the dynamic model lets PCr deplete
  continuously for as long as the phosphorylation state demands it — so more of
  the sustained sprint power is booked as alactic and less as lactic.

The upshot: a mechanistic, ADP/pH-regulated dynamic model, driven by the same
di Prampero demand and with only VLamax tuned, reproduces the **magnitude** of
the sprint's energy expenditure and its **aerobic** share, but partitions the
**anaerobic** energy along different lines than the empirical sprint model. The
two are best read as complementary views: the sprint model gives a compact,
fittable summary of the pathway peaks and time constants, while the dynamic model
adds an internally-consistent account of the phosphate and lactate dynamics that
produce them.

## The building blocks directly

Under the hood, `compare_bioenergetic_models()` builds the di Prampero demand for
the sprint, tunes VLamax with `tune_dynamic_vlamax()`, and integrates the model
with `simulate_dynamic_bioenergetics()`. We can call those exported building
blocks ourselves. Here we rebuild the demand for the trimmed sprint window with
the exported sprint-cost helpers, then drive the dynamic model directly.

```{r direct}
# extract and trim the sprint exactly as the sprint pipeline does
one <- gpexe[sprints$start_index[1]:sprints$end_index[1], ]
one <- trim_sprint_deceleration(trim_sprint_launch(one))

dt <- stats::median(diff(one$time))
demand <- cost_running_sprint(one$acc, one$velocity) * one$velocity
time <- one$time - min(one$time)

# tune VLamax to this demand (VO2max fixed), then simulate
tuned <- tune_dynamic_vlamax(demand, time, vo2max = vo2max)
sim <- simulate_dynamic_bioenergetics(demand, time, vo2max = vo2max,
                                      vlamax = tuned$vlamax)

c(vlamax = tuned$vlamax, rmse = tuned$rmse)
attr(sim, "energy_total")
head(sim)
```

`simulate_dynamic_bioenergetics()` returns the instantaneous partition (`alactic`,
`lactic`, `aerobic`, `total`, all W/kg) alongside the observed `measured` demand,
with the integrated energies (J/kg) attached as attributes — the same quantities
the comparison table aggregates.

