---
title: "Sprint Motion Analysis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Sprint Motion Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(runrgetics)
library(dplyr)
library(ggplot2)
```

# Sprint Motion Analysis

Ajuster la taille de sortie des graphiques....

The goal of this vignette is to provide a brief overview on the use of `runergetics` to analyze sprint motion based on Graubner and Nixdorf (2011) following the methodology detailed in a recent paper by Briand et al. (2025). (properly cite the package).

## Sprint Data

The package already includes s dataset `graubner_nixdorf_sprint` that contains the sprint data for the men's and women's 100, 200 and 400 m events at the Berlin 2009 World athletics championships. The same data sere used in the analysis by Briand et al. (2025) to propose a novel bioenergetic model of sprint running.

```{r}
head(graubner_nixdorf_sprints)
```

Data can be filtered depending on the event of interest. For example, to extract the data for Bolt's 100m. 

```{r}
bolt_100m <- graubner_nixdorf_sprints |>
  dplyr::filter(
    athlete == "Bolt",
    event == "Men's 100 m"
  )

bolt_100m
```

## Sprint Motion Analysis

From the data by Grabner and Nixdrof (2011) present time-splits recorded over different distance intervals for the different sprint events. di Prampero et al. (2005, 2018) have proposed a method to estimate the instantaneous cost of running based on velocity and acceleration data. This method is useful to evaluate the energy cost of accelerated running by making an analogy between accelerated running and running up an "Equivalent-Slope" and using the energy cost of uphill as proposed by Minetti (2002). 
From the energy cost of running the metabolic power can be computed as the product of the energy cost and running velocity. 
To apply this method, we need to estimate the instantaneous velocity and acceleration of the runner using the avalailable information from Graubner and Nixdorf (2011). A simple model to obtain the instanatenous vlocity and acceleration during a sprint has been described by di Prampero et al. (2018) and Briand et al. (2025). The function `sprint_motion_model_data` can be used to extract intantaneous distance, velocity, acceleration and metabolic power from `graubner_nixdorf_sprints` data. the following code chunk shows how to extract the data for Bolt's 100m event.

The `sprint_motion_model_data` function allows to adjust the parameter corresponding to the running energy cost on flat terrain. In the original equation proposed by Minetti et al. (2002), the cost of running on flat terrain shows an interindivual variability and different values have been reported over the years. di Prampero et al. (1986) reported slightly higher values of 3.8 J/kg/m and up to 3.86 J/kg/m. It was also suggested that for running at higher intensity, the energy cost was slightly increasd, up to around 3.8 J/kg/m. To remain consistent with these values, the default value for the cost of running of flat ground in Briand et al. (2025)'s article was set at 3.8 J/kg/m. To recover values similar to those presented in Briand et al.'s (2025) paper, 3.8 J/kg/m have to be set. However, using the default value of 3.6 J/kg/m will yield similar result results and will not affect the analysis of the general conceptualisation of the model. Along these lines, a small demonstration will be made in the `sprint_bioenergetic_model` vignette. 

Also, the method to compute the "Equivalent-Slope" of the acceleration can be adjusted. The `original` method is based on the energy cost of constant running velocity on different slopes proposed by Minetti et al. (2002). The `extended` method is based on the energy cost of running on extended slopes proposed by Minetti and Pavei (2018). The default value is set to `original` in the function.


```{r}
bolt_100m_motion_data <- sprint_motion_model_data(
  mean_velocity_splits = bolt_100m$velocity,
  time_splits = bolt_100m$splits,
  distance = bolt_100m$distance,
  reaction_time = bolt_100m$reaction_time[1],
  maximal_velocity = bolt_100m$maximal_velocity[1],
  cost_running_flat = 3.8, 
  slope_equation = "original"
)

head(bolt_100m_motion_data)
```

The function can be applied to any other event in the `graubner_nixdorf_sprints` dataset and to any external data sets that follow the same structure.

The sprint analysis can be ploted. The following code chunk allows to visualize Bolt's 100m event modeled distance, velocity, acceleration and metabolic power, along with the actual data provided in the `graubner_nixdorf_sprints` dataset.

```{r}
plot_sprint_motion_analysis(
  mean_velocity_splits = bolt_100m$velocity,
  time_splits = bolt_100m$splits,
  distance = bolt_100m$distance,
  reaction_time = bolt_100m$reaction_time[1],
  maximal_velocity = bolt_100m$maximal_velocity[1],
  cost_running_flat = 3.8
)
```

The 100 m events in the Graubner and Nixdorf are the only events where a maximal velocity is recorded. For the 200 and 400 m events, the maximal velocity is not available in the `graubner_nixdorf_sprints` dataset. It is referred to an `NA`. It can be left empty in the function call, and the function will estimate the maximal velocity based on the available data. 

Here is an exemple of the sprint motion data and plot analysis for the Men's 400 m.

```{r}
# filter Merrit's performance
merritt_400m <- graubner_nixdorf_sprints |>
  filter(
    athlete == "Merritt",
    event == "Men's 400 m"
  )

# extract sprint motion data
merritt_400m_motion_data <- sprint_motion_model_data(
  mean_velocity_splits = merritt_400m$velocity,
  time_splits = merritt_400m$splits,
  distance = merritt_400m$distance,
  reaction_time = merritt_400m$reaction_time[1],
  cost_running_flat = 3.8
)

# Visualize sprint motion analysis

plot_sprint_motion_analysis(
  mean_velocity_splits = merritt_400m$velocity,
  time_splits = merritt_400m$splits,
  distance = merritt_400m$distance,
  reaction_time = merritt_400m$reaction_time[1],
  cost_running_flat = 3.8
)
```
## Instantaneous Velocity and Power Sex Comparison

In our article (Briand et al. 2025) proposing a novel bioenergetic model of sprint running, we generated plots comparing the instantaneous velocity and power for men's and women's 100, 200 and 400 m events, along with the velocity data from the Graubner and Nixdorf dataset. The functions `sex_comparison_plot_velocity` and `sex_comparison_plot_power` can be used to generate these plots.

```{r}
# Sex comparison plots for the 100 m

# Extract the data for the 100 m
men_100 <- graubner_nixdorf_sprints |>
  filter(event == "Men's 100 m")

women_100 <- graubner_nixdorf_sprints |>
  filter(event == "Women's 100 m")

# Get the sprint motion data for both men and women

sprint_data_men_100 <- sprint_motion_model_data(
  mean_velocity_splits = men_100$velocity,
  time_splits = men_100$splits,
  distance = men_100$distance,
  reaction_time = men_100$reaction_time[1],
  maximal_velocity = men_100$maximal_velocity[1],
  cost_running_flat = 3.8
)

sprint_data_women_100 <- sprint_motion_model_data(
  mean_velocity_splits = women_100$velocity,
  time_splits = women_100$splits,
  distance = women_100$distance,
  reaction_time = women_100$reaction_time[1],
  maximal_velocity = women_100$maximal_velocity[1],
  cost_running_flat = 3.8
)


# Generate the plots

sex_comparison_plot_velocity(
  sprint_data_male = sprint_data_men_100,
  sprint_data_female = sprint_data_women_100,
  observed_data_male = men_100,
  observed_data_female = women_100,
  color = "darkgreen"
) + ggplot2::ggtitle("Sex Comparison of the 100 m Instantaneous Velocity")

sex_comparison_plot_power(
  sprint_data_male = sprint_data_men_100,
  sprint_data_female = sprint_data_women_100
) + ggplot2::ggtitle("Sex Comparison of the 100 m Instantaneous Metabolic Power")
```

The round points and solid line corresponds to the men's event, while the triangle points and dashed lines correspond to the women's event.

The same approach can be repeated for the 200 and 400 m events, or any other events as long as data used have the correct structure.

Now that we can analyze the sprint motion and the instanatneous metabolic power, we can move to the development of the sprint bioenergetic model to quantify instantaneous metabolic contributions from the anaeobic alactic, lactic and aerobic metabolisms.

### Testing the acceleration phase fit assumptions

A non-linear least-squared fit approach performed on the acceleration phase assuming an exponential rise of velocity to determine $\tau$. We want to check if the model's assumptions are met by looking at homoscedasticity and normality of the residuals.

The following function returns the velocity model for distances of less than 200 m. 


Now we can apply the model on the male and female 100, 200 and 400 m and check if the assumptions are respected. 

```{r}
time <- find_time_velocity(men_100$splits, men_100$reaction_time[1])
velocity <- men_100$velocity
reaction_time <- men_100$reaction_time[1]


# generate the model
mod_100 <- sprint_velocity_least_squares_model(time = time, velocity = velocity, reaction_time = reaction_time)
mod_100
```


A diagnostic of the non linear least-squares model assumptions can be made with the function `diagnose_velocity_model_acc`. The function will return the RMSE and RSE of the model, along with 4 plots: observed vs. fitted values, residuals vs. fitted values, Q–Q plot of residuals, and histogram of residuals.

```{r}
diagnose_velocity_model_acc(
  mean_velocity_splits = men_100$velocity,
  time_splits = men_100$splits,
  distance = men_100$distance,
  reaction_time = men_100$reaction_time[1]
)
```

As specified in Briand et al. (2025), for running distances beyond 200 m, the non-linear least-squares model uses only estimated instantaneous velocity and their associated time points up to distances of 200 m. In such case, the distance parameter can be modulated in the `diagnose_velocity_model_acc` function. 

We can apply the same function on Merritt's 400 m data and the fit will automatically consider only distances up to 200 m if the sprint distance of 400 m is specified. 

```{r}
diagnose_velocity_model_acc(
  mean_velocity_splits = merritt_400m$velocity,
  time_splits = merritt_400m$splits,
  distance = merritt_400m$distance,
  reaction_time = merritt_400m$reaction_time[1],
  sprint_total_distance = 400
)
```

We can apply a similar diagnostic for the whole sprint motion model using the function `diagnose_sprint_model`. This function will return the RMSE and RSE of the model, along with 4 plots: observed vs. fitted values, residuals vs. fitted values, Q–Q plot of residuals, and histogram of residuals.

```{r}
diagnose_sprint_model(
  mean_velocity_splits = men_100$velocity,
  time_splits = men_100$splits,
  distance = men_100$distance,
  reaction_time = men_100$reaction_time[1],
  maximal_velocity = men_100$maximal_velocity[1],
  cost_running_flat = 3.8
)
```

The same plot can be generated for each event. In this case, it is generated for the 400 m. 

```{r}
diagnose_sprint_model(
  mean_velocity_splits = merritt_400m$velocity,
  time_splits = merritt_400m$splits,
  distance = merritt_400m$distance,
  reaction_time = merritt_400m$reaction_time[1],
  maximal_velocity = merritt_400m$maximal_velocity[1],
  cost_running_flat = 3.8
)
```


## References


