---
title: "Sprint Lactic and Alactic Capacities"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Sprint Lactic and Alactic Capacities}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(runrgetics)
library(ggplot2)
library(dplyr)
```

The current vignette explores the modelling of the determination of lactic and alactic capacities from sprint performances. 

## A Model of Lactic and Alactic Energy Contributions as a Function of Race Duration

The first step consist of proposing a model for lactic and alactic energy contribution as a function of race duration. Briand et al. (2025) proposed such model which was determined by using Graubner and Nixdorf's (2011) dataset. The performances of the men's 100, 200 and 400 m were assumed to correspond to an hypothetical athlete's performances. Lactic and alactic energy contributions as a function of race duration were computed using the sprint bioenergetic model. 

The following functions `sprint_alactic_energy_duration` and `sprint_lactic_energy_duration` compute the alactic and lactic energy expenditure and average power for a given sprint event. The functions input is a tibble with a time column and instantaneous alactic and lactic power over the sprint, as computed using the sprint bioenergetic model proposed by Briand et al. (2025). 

The following chunk shows how to use the function with Bolt's 100 m performance data from the Graubner and Nixdorf dataset.

```{r}
# Extract Bolt's 100 m data from Graubner and Nixdorf data set.

# Extract Bolt's 100 m data from Graubner and Nixdorf data set.
bolt_100m <- graubner_nixdorf_sprints |>
  dplyr::filter(
    athlete == "Bolt",
    event == "Men's 100 m"
  )

# Compute sprint motion data

bolt_100m_motion_data <- sprint_motion_model_data(
  mean_velocity_splits = bolt_100m$velocity,
  time_splits = bolt_100m$splits,
  distance = bolt_100m$distance,
  reaction_time = bolt_100m$reaction_time[1],
  maximal_velocity = bolt_100m$maximal_velocity[1]
)

# Compute sprint power data using Briand et al.'s (2025) sprint bioenergetic model

bolt_modeled_data <- sprint_bioenergetic_model_data(bolt_100m_motion_data)

# Compute alactic energy vs duration for Bolt's 100 m performance

alactic_energy_duration_bolt_100m <- sprint_alactic_energy_duration(bolt_modeled_data)
alactic_energy_duration_bolt_100m

# Compute lactic energy vs duration for Bolt's 100 m performance

lactic_energy_duration_bolt_100m <- sprint_lactic_energy_duration(bolt_modeled_data)
lactic_energy_duration_bolt_100m
```

To formalise a model of lactic and alactic energy as a function of race duration, Briand et al. (2025) used the men's 100, 200 and 400 m performances from the Graubner and Nixdorf dataset. The function `sprint_alactic_energy_duration_graubner_nixdorf` and `sprint_lactic_energy_duration_graubner_nixdorf` can extract alactic and lacic energis as a function of race duraion for these events. 

```{r}
# extract men alactic energy vs duration for the 100, 200 and 400 m

men_alactic_energy <- sprint_alactic_energy_duration_graubner_nixdorf()
men_alactic_energy

# extract men lactic energy vs duration for the 100, 200 and 400 m

men_lactic_energy <- sprint_lactic_energy_duration_graubner_nixdorf()
men_lactic_energy
```

### Formulating a Model of Alactic Average Power as a Function of Race Duration

Based on the hypothetical male alactic and lactic energy and average power over race durations, two seperate models were proposed by Briand et al. (2025). These models enable the estimation of maximal alactic and lactic capacities as a function of race duration. 

The first model relates average alactic power $P_{al,avg}$ with race duration $T$. As described by Briand et al. (2025), a log-normal function can be used to model this functional relationship. The model was parametized to minimize the error between the average alactic power over the men's 100, 200 and 400 m, while keeping a good balance with theoretical behavior of alactic metabolism and available experimental data in the scientific litterature. More details are provided in the original article (Briand et al. 2025). 

The model follows the following equation: 

$$
P_{al,avg}(T) =  \frac{E_{al,max}}{T} \cdot \exp\left(\frac{-(ln(T)-1.75)^2}{2\cdot(1.5)^2}\right)
$$
Where $E_{al,max}$ corresponds to the alactic capacity. The function `sprint_alactic_duration_model` can be used to compute average alactic power as a function of race duration and maximal alactic capacity.

Although few data points are available, the average alactic power over the 100, 200 and 400 m can be used to estimate the hypothetical male athlete maximal alactic capacity through non-linear least squares fitting. 

The function `sprint_alactic_duration_model_fit` performs the non-linear least-squares fit. The function `sprint_alactic_duration_model_fit_rse` provides the fit residual standard error as provided by Briand et al. (2025) and the function `sprint_alactic_capacity` computes the maximal alactic capacity. 

```{r}
# Fit alactic model to power
sprint_alactic_duration_model_fit(men_alactic_energy)

# compute residual standard error
sprint_alactic_duration_model_fit_rse(men_alactic_energy)

# compute alactic capacity

sprint_alactic_capacity(men_alactic_energy)
```

The function `plot_sprint_alactic_duration` can be used to plot the alactic model fit with the average alactic power over the 100, 200 and 400 m.

```{r}

plot_sprint_alactic_duration(men_alactic_energy)

```

The following function `plot_sprint_alactic_duration_briand_article()` reproduces the plot presented in Briand et al. (2025) showing the alactic model fit with the average alactic power over the 100, 200 and 400 m, superposing the model applied to male and female data.

```{r}
plot_sprint_alactic_duration_briand_article()
```


### Formulating a Model of Alactic Energy as a Function of Race Duration

A second model can be used to model lactic energy $E_{la}$ as a function of race duration $T$. As described by Briand et al. (2025), a bi-exponential function can be used to model this functional relationship. The model was parametized to minimize the error between the lactic energy over the men's 100, 200 and 400 m, while keeping a good balance with theoretical behavior of lactic metabolism and available experimental data in the scientific litterature. More details are provided in the original article (Briand et al. 2025). This will also demonstrated in subsequent sections showing how to model fits data accumulated lactate over different running durations.

The lactic model as a function of race duration follows the following equation: 

$$
E_{la}(T) = E_{la,max} \cdot k_{norm,2} \left(1 - \exp\left(-\frac{(T-3)}{20}\right)\right) \cdot \exp\left(-\frac{(T-3)}{2000}\right)
$$

Where $E_{la,max}$ corresponds to the lactic capacity. $k_{norm_2}$ is a normalisation constant ensure the maximal lactic energy expenditure corresponds to the maximal lactic capacity. The function `sprint_lactic_duration_model` can be used to compute average alactic power as a function of race duration and maximal alactic capacity.

Although few data points are available, the average alactic power over the 100, 200 and 400 m can be used to estimate the hypothetical male athlete maximal alactic capacity through non-linear least squares fitting. 

The function `sprint_lactic_duration_model_fit` performs the non-linear least-squares fit. The function `sprint_lactic_duration_model_fit_rse` provides the fit residual standard error as provided by Briand et al. (2025) and the function `sprint_lactic_capacity` computes the maximal alactic capacity. 


```{r}
# Fit lactic model to power
sprint_lactic_duration_model_fit(men_lactic_energy)

# compute residual standard error
sprint_lactic_duration_model_fit_rse(men_lactic_energy)

# compute lactic capacity

sprint_lactic_capacity(men_lactic_energy)
```

The function `plot_sprint_lactic_duration_model` can be used to plot the lactic model fit with the average lactic energy expenditure over the 100, 200 and 400 m.


```{r}
plot_sprint_lactic_duration(men_lactic_energy)
                            
```


### Accumulated Lactate and Kindemann (1977) data

Kindermann's (1977) provides data of blood lactate concentration measured on maximal efforts of different durations. The data is provided in the `kindermann_lactate` dataset. 

```{r}
head(kindermann_lactate)
```

Accumulated lactate corresponds to the blood lactate accumulated above resting level concentrations and reflects the expenditure of lactic energy. The function `convert_acc_lactate_to_lactic_energy` can be used to convert accumulated lactate to lactic energy expenditure in J/kg. Alternativeyl, the function `convert_lactic_energy_to_acc_lactate` can be used to convert lactic energy expenditure in J/kg to accumulated lactate in mmol/L.
From these functions, we can estimate lactic energy from Kindermann's (1977) accumulated blood lacate data.

```{r}
kindermann_lactic_energy <- kindermann_lactate |> 
  mutate(lactic_energy = convert_acc_lactate_to_lactic_energy(accumulated_lactate))

head(kindermann_lactic_energy)
```

In Briand et al. (2025), the lactic energy model was fitted to the Kindermann's (1977) data. The function `plot_sprint_lactic_duration_kindermann` can be used to generate the Figure presented in Briand et al. (2025) article.

```{r}
plot_sprint_lactic_duration_kindermann()
```

The fit seems to everestimate the lactic energy expenditude for longer running duration. With a faster decay, the model seems to better fit the estimated lactioc energy using Kindermann's (1977) data. 

```{r}
plot_sprint_lactic_duration_kindermann(k1 = 20, k2 = 1500)
```

As mentionned by Briand et al. (2025), further work is needed to asses the shape of the lactic energy model for longer running durations. Kindermann's data represent a way to assess that the biexponential model to map lactic energy expenditure as a function of running duration is a good avenue that needs refinement. Further data are needed and some of the limits of Kindermann's (1977) data to address this challenge are discussed in Briand et al. (2025).
