---
title: "Sprint Bioenergetic Model"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Sprint Bioenergetic Model}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(runrgetics)
library(ggplot2)
library(dplyr)
```

The previous vignettes `sprint energetics exploration` and `sprint model formalisation` explore the functions that can represent the energy pathways in sprints and the set of parameters that can be chosen such that the functions apply on all three distances for men and women sprint performances. 

This vignette assembles the three functions into a bioenergetic model of sprint that can be used to quantify the instantaneoues energy contribution of the three energy pathways during a sprint. The model is described in details in Briand et al. 2025.

We can use the male 100 m sprint data from the Graubner and Nixdorf (2023) dataset to illustrate the possible applications of the sprint bioenergetic model.

Supposing that the maximal alactic power attained during the sprint is of 150 W/kg and that the maximal lactic power attained during the sprint is of 50 W/kg, we can use the function `sprint_bioenergetic_model` to compute the total metabolic power during the sprint.

```{r}
# Setting, time, maximal_alactic_power and maximal_lactic_power
time <- seq(0, 9.58, by = 0.01)
maximal_alactic_power <- 150
maximal_lactic_power <- 50

# applying the sprint_bioenergetic_model function to get total metabolic power

total_power <- sprint_bioenergetic_model(time, maximal_alactic_power, maximal_lactic_power)

# plot total power as a function of time

ggplot2::ggplot(data = tibble::tibble(time = time, total_power = total_power), aes(x = time, y = total_power)) +
  ggplot2::geom_line(linewidth = 1.5) +
  ggplot2::labs(
    title = "Total Metabolic Power during a Sprint",
    x = "Time (s)",
    y = "Total Power (W/kg)"
  ) +
  ggplot2::theme_minimal()
```

The `output` argument of the function `sprint_bioenergetic_model` allows to compute the contribution of each energy pathway to the total metabolic power. For example, we can compute the alactic power contribution during the sprint as follows:

```{r}
# applying the sprint_bioenergetic_model function to get total metabolic power

alactic_power <- sprint_bioenergetic_model(time, maximal_alactic_power, maximal_lactic_power, output = "alactic power")

# plot total power as a function of time

ggplot2::ggplot(data = tibble::tibble(time = time, alactic_power = alactic_power), aes(x = time, y = alactic_power)) +
  ggplot2::geom_line(color = "darkred", linewidth = 1.5) +
  ggplot2::labs(
    title = "Alactic Power during a Sprint",
    x = "Time (s)",
    y = "Alactic Power (W/kg)"
  ) +
  ggplot2::theme_minimal()
```


## Extracting Sprint Maximal Alactic and Maximal Lactic from Sprint Motion Data

The previous examples relied on rough guesses of the sprint maximal alactic and maximal lactic power. The function `fit_sprint_bioenergetic_model` can be used to fit the sprint bioenergetic model to the sprint motion data and extract the maximal alactic and maximal lactic power.

The following example shows how to derive sprint motion data from the Graubner and Nixdorf (2011) dataset and fit the sprint bioenergetic model to the data to extract the sprint maximal alactic and maximal lactic power. The functions `sprint_bioenergetic_model_max_al` and `sprint_bioenergetic_model_max_la` are used to extract both paramters.

```{r}
# Extract Bolt's 100 m data from Graubner and Nixdorf data set.
bolt_100m <- graubner_nixdorf_sprints |>
  dplyr::filter(
    athlete == "Bolt",
    event == "Men's 100 m"
  )

# Compute sprint motion data

bolt_100m_motion_data <- sprint_motion_model_data(
  mean_velocity_splits = bolt_100m$velocity,
  time_splits = bolt_100m$splits,
  distance = bolt_100m$distance,
  reaction_time = bolt_100m$reaction_time[1],
  maximal_velocity = bolt_100m$maximal_velocity[1], 
  cost_running_flat = 3.8
)

# extract maximal alactic and maximal lactic powers

maximal_alactic_power <- sprint_bioenergetic_model_max_al(bolt_100m_motion_data)
maximal_alactic_power

maximal_lactic_power <- sprint_bioenergetic_model_max_la(bolt_100m_motion_data)
maximal_lactic_power
```

### Cost of running

As mentioned in the `sprint_motion_analysis` vignette, to recover similar values as the ones presented in Briand et al. (2025) article, the cost of running on flat ground has to be set to 3.8 J/kg/m. The choice of the value was made to be consistent with the different values of energy cost of running on flat ground in the litterature. The default value of 3.6 J/kg/m can also be used. This value was proposed by Minetti et al. (2002) and Minetti and Pavei (2018).

We can compare the energy expended using a cost of running on falt terrain of 3.8 J/kg/m and 3.6 J/kg/m.

```{r}

# compute with cost of running on flat terrain of 3.6 J/kg/m

bolt_motion_data_3_6 <- sprint_motion_model_data(
  mean_velocity_splits = bolt_100m$velocity,
  time_splits = bolt_100m$splits,
  distance = bolt_100m$distance,
  reaction_time = bolt_100m$reaction_time[1],
  maximal_velocity = bolt_100m$maximal_velocity[1],
  cost_running_flat = 3.6, 
  dt = 0.01
)

# compute max power and total energy exepnded

max_3_8 <- sprint_maximum_metabolic_power(bolt_100m_motion_data)
max_3_6 <- sprint_maximum_metabolic_power(bolt_motion_data_3_6)

max_3_8 - max_3_6



# total energy expended

total_energy_3_8 <- energy_total(bolt_100m_motion_data, type = "power")
total_energy_3_6 <- energy_total(bolt_motion_data_3_6, type = "power")

total_energy_3_8 - total_energy_3_6


```

With Bolt data, we obtained the same maximum, because for high Equivalent Slope, the correction proposed by di Prampero and Osgnach (2018) leads to the same result. The energy difference is around 20 J/kg, which corresponds to approximately a 2.5 % difference over the total energy expenditure. 

For high acceleration, the method proposed by Minette and Pavei (2018) can be used since they used data from Giovanelli et al. (2016), that measured the energy cost of running on greater slopes than the ones sutided by Minetti et al. (2002)

```{r}

# compute cost of running using Minetti and Pavei extended equation

bolt_motion_data_extended <- sprint_motion_model_data(
  mean_velocity_splits = bolt_100m$velocity,
  time_splits = bolt_100m$splits,
  distance = bolt_100m$distance,
  reaction_time = bolt_100m$reaction_time[1],
  maximal_velocity = bolt_100m$maximal_velocity[1],
  slope_equation = "extended"
)

max_ext <- sprint_maximum_metabolic_power(bolt_motion_data_extended)
max_3_6 <- sprint_maximum_metabolic_power(bolt_motion_data_3_6)

max_ext - max_3_6



# total energy expended

total_energy_ext <- energy_total(bolt_motion_data_extended, type = "power")
total_energy_3_6 <- energy_total(bolt_motion_data_3_6, type = "power")

total_energy_ext - total_energy_3_6

```

There is a small difference of 4.73 J/kg in energy expended, which represents a 0.6 % difference over the total energy expenditure. The maximum metabolic power is also slightly higher with the original equation, this suggests that the maximal power could have been slighlty overestimated with the original method extrpolating energy cost for greater slopes. 

Briand et al. (2025) compared the results obtained from both approaches and discuss their implications. While the differences don't affect the bioenergetic model application and its utility, it is important to be aware of the differences in the energy expenditure estimates when using different cost of running equations. It is also worth understanding the limits of the original approach.

## Extracting Sprint Maximal Alactic and Maximal Lactic from Women's Sprint Motion Data

As presented in Briand et al. (2025), for elite sprinters available in the Graubner and Nixdrof (2011) dataset, the maximal aerobic power can be estimated to be 24.5 W/kg for male sprinters and 21 W/kg for female sprinter. The default parameter was set at 24.5 W/kg for the maximal aerobic power. When dealing with female athletes' performances, the maximal aerobic power can be adjusted accordingly. 

The following example derived Fraser-Pryce's 100 m maximal alactic and maximal lactic power from the Graubner and Nixdorf (2011) dataset.

```{r}
# Extract Bolt's 100 m data from Graubner and Nixdorf data set.
FP_100m <- graubner_nixdorf_sprints |>
  dplyr::filter(
    event == "Women's 100 m"
  )

# Compute sprint motion data

FP_100m_motion_data <- sprint_motion_model_data(
  mean_velocity_splits = FP_100m$velocity,
  time_splits = FP_100m$splits,
  distance = FP_100m$distance,
  reaction_time = FP_100m$reaction_time[1],
  maximal_velocity = FP_100m$maximal_velocity[1], 
  cost_running_flat = 3.8
)

# extract maximal alactic and maximal lactic powers

maximal_alactic_power <- sprint_bioenergetic_model_max_al(FP_100m_motion_data, maximal_aerobic_power = 21)
maximal_alactic_power

maximal_lactic_power <- sprint_bioenergetic_model_max_la(FP_100m_motion_data, maximal_aerobic_power = 21)
maximal_lactic_power
```

From the sprint motion data extracted from the Graubner and Nixdorf data set, the function `sprint_bioenergetic_model_data` can be used to add columns to the sprint motion data, representing the energy from the 3 energy pathways as well as the total metabolic power computed from the model. 

```{r}
bolt_modeled_data <- sprint_bioenergetic_model_data(bolt_100m_motion_data)

head(bolt_modeled_data)
```


## Visualising the Sprint Bioenergetic Model Compared to Sprint Motion Data

The function `plot_sprint_bioenergetic_model` can be used to visualise the power derived sprint bioenergetic model compared to the metabolic power available in the sprint motion data, derived using di Pampero et al. (2005, 2018) approach. The following example plots the sprint bioenergetic model compared to the sprint motion data for Bolt's 100 m sprint. 

```{r}
plot_sprint_bioenergetic_model(
  sprint_motion_data = bolt_100m_motion_data
)
```

The same function can be used to plot the sprint bioenergetic model compared to the sprint motion data for Fraser-Pryce's 100 m sprint, specifying a maximal aerobic power of 21 W/kg. The y-axis scale can also be adjusted using the `scale_max_power` argument, which is set by default at 160 W/kg.

```{r}
plot_sprint_bioenergetic_model(
  sprint_motion_data = FP_100m_motion_data,
  maximal_aerobic_power = 21,
  scale_max_power = 110
)
```

## Sprint Bionergetic Model Goodness of Fit

The $R^2$, adjusted $R^2$ residual standard error (RSE) and root mean square error (RMSE) can be used to assess the goodness of fit of the sprint bioenergetic model. A function to evaluate each goodness of fot metric is available. 

```{r}
# residual standard error (rse)
sprint_bioenergetic_model_rse(
  sprint_motion_data = bolt_100m_motion_data,
  maximal_aerobic_power = 24.5
)


# root mean square error (rmse)

sprint_bioenergetic_model_rmse(
  sprint_motion_data = bolt_100m_motion_data,
  maximal_aerobic_power = 24.5
)

# R squared

sprint_bioenergetic_model_R2(
  sprint_motion_data = bolt_100m_motion_data,
  maximal_aerobic_power = 24.5
)

# Adjusted R squared

sprint_bioenergetic_model_adj_R2(
  sprint_motion_data = bolt_100m_motion_data,
  maximal_aerobic_power = 24.5
)
```


The function `sprint_bioenergetic_model_diagnostics` can be used to comute the metrics and visualize the residuals distribution and Q-Q plots to assess non-linear least-squares fit assumptions. The following example shows how to use the function to compute the goodness of fit for Bolt's 100 m sprint.

```{r}
sprint_bioenergetic_model_diagnostic(
  sprint_motion_data = bolt_100m_motion_data,
  maximal_aerobic_power = 24.5,
  histogram_binwidth = 1,
)
```

The fit has several issues and some patterns can be observed in the residuals. These patterns are likely due to the the fact that the metabolic power computed from the sprint motion data relies on an imperfect model of the sprint motion, where instantaneous velocity is reconstructed to the best of our ability with limited information, i.e. data points collected at 10, 20 or even 50 m intervals. A better model of the sprint motion and a more precise measurement of the instantaneous velocity would likely improve the fit of the sprint bioenergetic model to the sprint motion data.

The general bioenergetic model of sprunt running is also based on general functions that fit reasonnable well the approximate power distribution from the three metabolic pathways. The parameters' values were adjusted to ensure a good fit amongst all men and women running distances (see `sprint_model_formalisation` vignette). Therefore, while the model can be used to estimate energy contribution, it doesn't perfectly fit the metabolic power computed using the velocity, acceleration and equivalent slope as proposed by di Prampero et al. (2005, 2018).

We can try to specifically retrieve the parameters that best fit Bolt's 100 m to see if the diagnostic of the sprint bioenergetic model is better. 

```{r}
# get sprint optimal alactic power parameters

alactic_params <- sprint_alactic_model_optimal_params() |>
  filter(event == "Men's 100 m")

# extract optimal mu and sigma

mu <- alactic_params$mu
sigma <- alactic_params$sigma


# get sprint optimal lactic power parameters

lactic_params <- sprint_lactic_model_optimal_params() |>
  filter(event == "Men's 100 m")

k1 <- lactic_params$k1
k2 <- lactic_params$k2

# rerun the diagnostic with the updated parameters

sprint_bioenergetic_model_diagnostic(
  sprint_motion_data = bolt_100m_motion_data,
  mu = mu,
  sigma = sigma,
  k1 = k1,
  k2 = k2,
  maximal_aerobic_power = 24.5,
  histogram_binwidth = 1
)
```

We see that the diagnostic of the model is slightly improved. All the metrics (RSE, RMSE, $R^2$ and adjusted $R^2$) show improvements. The main discrepencies are observed in the early moment of the sprint, where the log-normal function of the alactic model doesn't perfectly match the metabolic power. The second discrepency is observed around metabolic eak power, where the alactic model lead to a slight overestimation of the peak power. The last discrepency is related to the transition between the acceleration and the constant-velocity phase, derived from the sprint motion model that reproduces the instantaneous velocity profile from the time splits. The limitations of the sprint bioenergetic model are discussed in further details by Briand et al. (2025).

To further evaluate the sprint bioenergetic model and its goodness of fit, we can compare the model's predicted energy expenditure from the three energy pathways with experimental data available in the literature.

## Sprint Bioenergetic Model Theortical Comparisons

### Total Energy Expenditure Computations and Comparisons

Using `bolt_modeled_data`, we can compute the total energy expenditure estimated using di Prampero et al. (2005, 2018) equivalent slope approach and compare it to the energy expenditure estimated using the sprint bioenergetic model. The following example shows how to compute the total energy expenditure for Bolt's 100 m sprint and compare it to the energy expenditure estimated using the sprint bioenergetic model.


```{r}
# energy estimated using di Prampero et al. (2005, 2018) equivalent slope approach
energy_estimated <- energy_total(bolt_modeled_data, type = "power")
round(energy_estimated)

# energy estimated using the sprint bioenergetic model

energy_bioenergetic_model <- energy_total(bolt_modeled_data, type = "power bioenergetic model")
round(energy_bioenergetic_model)

# we can compute the percentage difference between the two energy expenditure estimates

percentage_difference <- (energy_bioenergetic_model - energy_estimated) / energy_estimated * 100
round(percentage_difference, 2)
```


### Energy Pathways Expenditure Computations and Comparisons

Similarly, we can compute the energy exependiture from the three metabolic pathways estimated using the sprint bioenergetic model.

```{r}
# alactic energy expenditure

alactic_energy <- energy_alactic(bolt_modeled_data)
round(alactic_energy)

# lactic energy exepnditure
lactic_energy <- energy_lactic(bolt_modeled_data)
round(lactic_energy)

# aerobic energy expenditure
aerobic_energy <- energy_aerobic(bolt_modeled_data)
round(aerobic_energy)
```

Finally, we can compute the percentage of energy expenditure from each energy pathway relative to the total energy expenditure estimated using the sprint bioenergetic model.

```{r}
# alactic energy percentage
alactic_energy_percentage <- alactic_energy_percentage(bolt_modeled_data, type = "power bioenergetic model")
round(alactic_energy_percentage)

# lactic energy percentage

lactic_energy_percentage <- lactic_energy_percentage(bolt_modeled_data, type = "power bioenergetic model")
round(lactic_energy_percentage)

# aerobic energy percentage
aerobic_energy_percentage <- aerobic_energy_percentage(bolt_modeled_data, type = "power bioenergetic model")
round(aerobic_energy_percentage)

# anaerobic energy percentage
anaerobic_energy_percentage <- anaerobic_energy_percentage(bolt_modeled_data, type = "power bioenergetic model")
round(anaerobic_energy_percentage)
```

### Computing Maximal Metabolic Power

maximal metabolic power can be computed for the etabolic power derived using di Prampero et al. (2005, 2018) equivalent slope approach and for the metabolic power derived using the sprint bioenergetic model. The following example shows how to compute the maximal metabolic power for Bolt's 100 m sprint and compares the maximum obtained from the 2 approaches.

```{r}
# maximal metabolic power from di Prampero et al. (2005, 2018) equivalent slope approach

estimated_max_power <- sprint_maximum_metabolic_power(bolt_modeled_data, type = "power")
round(estimated_max_power)

# maximal metabolic power from the sprint bioenergetic model
estimated_max_power_bioenergetic_model <- sprint_maximum_metabolic_power(bolt_modeled_data, type = "power bioenergetic model")
round(estimated_max_power_bioenergetic_model)

# percentage difference between the two maximum metabolic power estimates

percentage_difference_max_power <- (estimated_max_power_bioenergetic_model - estimated_max_power) / estimated_max_power * 100
round(percentage_difference_max_power, 2)
```

### Distance Travelled Computation

The sprint bioenergetic model provides time series of the power produced by a sprinter over a sprint event. To validate the model's prediction it is relevent to invert the approach and compute distance travelled, as well as instantaneous velocity and acceleration from the power time series. 
The function `sprint_recover_distance` can be used to compute the distance travelled from a power time series. The following example shows how to use the function to recover the motion data from Bolt's 100 m sprint.

The Briand et al. (2025) paper mentions that the method to recover the distance travelled from the time series leads to a distance error of less than 0.01%, which can effectively be observed for time steps of 0.001 s. We can observe this in the following example.

```{r}
# Recompute Bolt motion data with a smaller time steps

bolt_motion_small_dt <- sprint_motion_model_data(
  mean_velocity_splits = bolt_100m$velocity,
  time_splits = bolt_100m$splits,
  distance = bolt_100m$distance,
  reaction_time = bolt_100m$reaction_time[1],
  maximal_velocity = bolt_100m$maximal_velocity[1],
  dt = 0.001, 
  cost_running_flat = 3.8
)

# Estimate distance through numerical integration of velocity

distance_actual <- pracma::trapz(bolt_motion_small_dt$time, bolt_motion_small_dt$velocity)

distance_recovered <- sprint_recover_distance(bolt_motion_small_dt$power, dt = 0.001, cost_running_flat = 3.8)

distance_percentage_diff <- (distance_recovered - distance_actual) / distance_actual * 100
round(distance_percentage_diff, 2)
```

Using such small time steps can be computationally heavy and therefore, for the remaining of the calculations, we will use a time step of 0.01 s, which is the default value for the `dt` argument in the `sprint_motion_model_data` function. This leads to slightly larger distance errors recovered from the power time series. The following code chunk also compute the distance difference as derived from di Prampero et al.'s (2005, 2018) equivalent slope approach and the distance recovered from the sprint bioenergetic model power time series. The percentage difference in Bolt's traveled distance over the 100 m is of less than 1 % (0.85 %). It's worth mentionning that the distance obtained from the `sprint_motion_model_data` function is not actually equal to distance traveled by Bolt which should be equal to 100 m. This occurs because the model to reconstruct instantaneous velocity from time splits relies on several assumptions that are discussed at length by Briand et al. (2025).

```{r}
# Recompute Bolt actual distance from motion data with a time step of 0.01 s

distance_actual <- pracma::trapz(bolt_100m_motion_data$time, bolt_100m_motion_data$velocity)

distance_recovered <- sprint_recover_distance(bolt_modeled_data$power, dt = 0.01, cost_running_flat = 3.8)

distance_percentage_diff <- (distance_recovered - distance_actual) / distance_actual * 100
round(distance_percentage_diff, 2)

# Compute the difference between modelled distance (recovered from power)

distance_recovered_bioenergetic_model <- sprint_recover_distance(bolt_modeled_data$power_mod, dt = 0.01, cost_running_flat = 3.8)

modelled_distance_percentage_diff <- (distance_recovered_bioenergetic_model - distance_recovered) / distance_recovered * 100
round(modelled_distance_percentage_diff, 2)

# A function was written to compute the percentage error between the modelled distance and the recovered distance from the power time series.

round(sprint_modeled_distance_percentage_error(bolt_modeled_data, dt = 0.01, cost_running_flat = 3.8), 2)
```

### Generating Comparison Tables

The previous computations can be used to generate the tables assessing the sprint bioenergetic model's goodness of fit and comparing the fit amongst distances in Briand et al. (2025) article. 

The goodness of fit metrics of each event can be computed using the function `sprint_bioenergetic_model_gof_metrics` that provides a data frame of the relevant metrics for the chosen event. 

```{r}
sprint_bioenergetic_model_gof_metrics(bolt_modeled_data, event = "Men's 100 m")
```

The following function returns a table similar to Table 2 in Briand et al. (2025) article, with the goodness of fit metrics for each event in the Graubner and Nixdorf (2011) dataset. Discrepencies between the table of the article can be due to the fact that the difference in distance travelled in this table was compared to the distance reconstructed from the estimated metabolic pouwer of sprint motion data, while in the article, the distance was compared to the distance traveled by the athlete as obtained from the instantaneous velocity model. Other small differences observed can be due to the choice of integration and integration steps.


```{r}
sprint_briand_article_gof_table()
```

The energy contribution from the 3 metabolic pathways of each event can be computed using the function `sprint_energy_contributions` that provides a data frame of the relevant energy contribution and percentage to total energy contribution for the chosen event.

```{r}
sprint_energy_contributions(bolt_modeled_data, event = "Men's 100 m")
```


The following function returns a table similar to Table 3 in Briand et al. (2025) article, with the energy contribution from the three metabolic pathways for each event in the Graubner and Nixdorf (2011) dataset. Discrepencies between the table of the one provided in the article can be due to the choice of integration and integration steps.


```{r}
sprint_energy_cont_briand_article_table()
```
Briand et al.'s (2025) article discuss the energy contribution of the metabolic pathways and puts the results in perspective with the literature. 


## Further cost of running comparison on longer sprint distances

The initial comparison was made on Bolt's 100 m, where the acceleration is strong. We can further test on Merrit's 400 m 

```{r}
# Extract Bolt's 100 m data from Graubner and Nixdorf data set.
men_400 <- graubner_nixdorf_sprints |>
  dplyr::filter(
    athlete == "Merritt",
    event == "Men's 400 m"
  )

# Compute sprint motion data

men_400_motion_data <- sprint_motion_model_data(
  mean_velocity_splits = men_400$velocity,
  time_splits = men_400$splits,
  distance = men_400$distance,
  reaction_time = men_400$reaction_time[1],
  maximal_velocity = men_400$maximal_velocity[1], 
  cost_running_flat = 3.8
)


# compute with cost of running on flat terrain of 3.6 J/kg/m

men_400_3_6 <- sprint_motion_model_data(
  mean_velocity_splits = men_400$velocity,
  time_splits = men_400$splits,
  distance = men_400$distance,
  reaction_time = men_400$reaction_time[1],
  maximal_velocity = men_400$maximal_velocity[1],
  cost_running_flat = 3.6, 
  dt = 0.01
)

# compute max power and total energy exepnded

max_3_8 <- sprint_maximum_metabolic_power(men_400_motion_data)
max_3_6 <- sprint_maximum_metabolic_power(men_400_3_6)

max_3_8 - max_3_6



# total energy expended

total_energy_3_8 <- energy_total(men_400_motion_data, type = "power")
total_energy_3_6 <- energy_total(men_400_3_6, type = "power")

total_energy_3_8 - total_energy_3_6

```

We observe a very small difference in maximal power. A difference is observed in total energy expenditure, which is due to the difference in the cost of running on flat terrain. The difference in cost shifts the power curve slightly, over longer duration, this leads to a difference in energy expenditure.

```{r}
men_400_motion_data$power[length(men_400_motion_data$power)] - men_400_3_6$power[length(men_400_3_6$power)] 
```

At the end of the 400 m , the difference in power output is less than 2 W/kg. As the power is only shifted, the bioenergetic model will apply with different energy cost values on flat terrain, just as it applies with women performances, who display smaller power output, while maintaining the overall shape of the power distribution. 

As a final exercise, we can plot the sprint bioenergetic model for Merrit's 400 m sprint, using the cost of running on flat terrain of 3.8 J/kg/m and 3.6 J/kg/m.

```{r}
plot_sprint_metabolic_power(mean_velocity_splits = men_400$velocity,
  time_splits = men_400$splits,
  distance = men_400$distance,
  reaction_time = men_400$reaction_time[1],
  maximal_velocity = men_400$maximal_velocity[1], 
  cost_running_flat = 3.8) + geom_line(data = men_400_3_6, aes(x = time, y = power), color = "darkred")
  
```

## Application of the bioenergtic model on modified energy cost equations

Finally, we can apply the bioenergtic model to this data (with energy cost of running of 3.6)

```{r}

plot_sprint_bioenergetic_model(
  sprint_motion_data = men_400_3_6
)

```

It fits the data reasonable good even with different cost of running on flat terrain. 

We can also test the application of sprint motion data obtained from Minetti and Pavei (2018) extended slope equation

```{r}
plot_sprint_bioenergetic_model(
  sprint_motion_data = bolt_motion_data_extended
)

```

Again, bioenergtic model appears to be reasonable good. Therefore, the choice of the energy cost of running on flat terrain has little impact on the model value and application. 
