---
title: "Sprint Training Analysis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Sprint Training Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4
)
```

```{r setup}
library(runrgetics)
library(dplyr)
library(ggplot2)
```

This vignette walks through the sprint-training analysis tools of `runrgetics`
end to end, using two track sessions that were recorded **simultaneously** by a
gpexe local-positioning unit and a Polar/Stryd watch. Starting from the raw paired
recordings we filter the noisier watch signal, compare the two devices, detect the
individual sprints of a workout, compare them, and finally decompose each sprint's
metabolic power into its aerobic, anaerobic-alactic and anaerobic-lactic
contributions following di Prampero (2005, 2018) and Briand et al. (2025).

## The paired datasets

The package ships two paired-device sessions as lazy-loaded datasets: a mixed
sprint session (`sprint_mix_paired`) and a session of ten 200 m sprints
(`ten_200_sprints_paired`). Each is a tidy long table with a `source` column
identifying the device, a synchronized `time` (s) and, where available, velocity,
position and power channels (see `?sprint_mix_paired`).

```{r}
head(ten_200_sprints_paired)

table(ten_200_sprints_paired$source)
```

The gpexe unit samples faster and applies its own onboard filtering, so it is
treated as the reference signal; the watch (`polar_stryd`) is noisier and is what
we filter below.

## Filtering the watch signal

`filter_watch_motion()` low-pass filters the watch data so that its speed,
acceleration and distance approximate the gpexe-filtered reference. By default it
applies a zero-phase Butterworth filter (cut-off 0.175 Hz), resamples to a uniform
grid and derives a clean acceleration and cumulative distance.

```{r}
watch <- ten_200_sprints_paired |> filter(source == "polar_stryd")

filtered_watch <- filter_watch_motion(watch)

head(filtered_watch)
```

## Comparing the devices

`compare_devices()` filters the watch, derives matched motion and power channels
for both devices on a common grid, and reports their agreement (root-mean-square
error, bias, correlation and magnitude ratio). gpexe is the reference, so a
`ratio` above 1 means the watch reads higher.

```{r}
compare_devices(ten_200_sprints_paired)
```

`plot_device_comparison()` overlays a chosen signal for the filtered watch and the
gpexe reference.

```{r}
plot_device_comparison(ten_200_sprints_paired, signal = "speed")
```

## Detecting sprints automatically

`detect_sprints()` segments a continuous workout into individual high-intensity
efforts. By default it flags spans where speed exceeds 5 m/s for at least 2 s,
merging efforts separated by short recoveries. Here it recovers the ten 200 m
repetitions of the workout.

```{r}
gpexe <- ten_200_sprints_paired |> filter(source == "gpexe")

sprints <- detect_sprints(gpexe)

sprints
```

Each effort comes with its start/end (time and row index), duration, peak and mean
speed, distance and peak acceleration.

## Comparing a multi-sprint workout

`compare_workout_sprints()` summarises every detected effort for side-by-side
comparison, and `plot_workout_sprints()` overlays each effort against distance from
its start (speed by default), reproducing the classic sprint-comparison figure.

```{r}
compare_workout_sprints(gpexe, sprints = sprints)
```

```{r}
plot_workout_sprints(gpexe, sprints = sprints)
```

The efforts are shown over their full profile, from the launch through the
deceleration, so the within-workout consistency (and any fatigue) is visible.

## Training bioenergetic analysis

Finally, `analyze_training_bioenergetics()` applies the sprint bioenergetic model
to each detected sprint and splits its metabolic power into alactic, lactic and
aerobic contributions. The maximal aerobic power (MAP) is a parameter; for these
two sessions we use **27 W/kg**. Each sprint's end-of-effort deceleration is
trimmed automatically before fitting, since the model is only valid up to the
sprint finish.

```{r}
bioenergetics <- analyze_training_bioenergetics(
  gpexe,
  sprints = sprints,
  maximal_aerobic_power = 27
)

bioenergetics$per_sprint

bioenergetics$summary
```

We can visualise the decomposition for a single sprint. We extract the effort
(with a short pre-roll to capture the launch), compute its metabolic power and pass
it to `plot_sprint_bioenergetic_model()`.

```{r}
sp <- sprints[2, ]
segment <- gpexe[max(1, sp$start_index - 15):sp$end_index, ]

segment_power <- compute_metabolic_running_power(segment[, c("time", "velocity")])

sprint_md <- tibble(
  time = segment_power$time - min(segment_power$time),
  power = segment_power$metabolic_power
)

plot_sprint_bioenergetic_model(sprint_md, maximal_aerobic_power = 27,
                               scale_max_power = 60)
```

Across the ten repetitions, the aerobic and lactic pathways dominate these ~30 s
efforts, with a smaller alactic contribution concentrated in the acceleration
phase — as expected for repeated 200 m sprints.

## References

- di Prampero, P. E., et al. (2005). Sprint running: a new energetic approach.
  *Journal of Experimental Biology*, 208(14), 2809–2816.
- Osgnach, C., & di Prampero, P. E. (2018). Metabolic power in team sports.
  *International Journal of Sports Medicine*.
- Minetti, A. E., et al. (2002). Energy cost of walking and running at extreme
  uphill and downhill slopes. *Journal of Applied Physiology*.
- Minetti, A. E., & Pavei, G. (2018). Update and extension of the 'equivalent
  slope' of speed-changing level locomotion. *Journal of Experimental Biology*.
- Briand, J., et al. (2025). *European Journal of Applied Physiology*.
  doi:10.1007/s00421-025-05831-0
