---
title: "Simulating Athlete Sprint Performances"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Simulating Athlete Sprint Performances}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(runrgetics)
library(dplyr)
```

The last part of Briand et al. (2025) article presents a simulation of fictitious male and female sprinters who could theoretically realsied respectiveley the men's and women's 100, 200 and 400 m performances in the `graubner_nixdorf_sprints` data set. The simulation is based on the sprint bioenergetic model as well as the alactic power-duration and lactic energy duration models.

## Finiding Maximal Lactic and Alactic Power in a Sprint From Capacities

The first step to achieve the simulation involves estimating what are the sprint's maximal alactic and lactic power, which are parameters of the sprint bioenergetic model, given the fictitious athlete's maximal alactic and lactic capacities. 

The function `find_max_al` finds the sprint maximal alactic power, given the sprint duration and the athlete alactic capacity. Similarly, the function `find_max_la` finds the sprint maximal lactic power, given the sprint duration and the athlete lactic capacity. 

```{r}

# set example duration and capacities
duration <- 20
lactic_capacity <- 1350
alactic_capacity <- 350

# find the maximal alactic and lactic power
find_max_al(duration, alactic_capacity)
find_max_la(duration, lactic_capacity)

```

## Run a Simulation to Estimate Fictitious Athletes Lactic and Alactic Capacities

For the upcoming section we consider that the men's 100, 200 and 400 m performances in the `graubner_nixdorf_sprints`dataset are performed by the same fictitious male athlete and similarly, that the women's 100, 200 and 400 m performances are performed by the same fictitious female athlete. 

The function `graubner_nixdorf_perf_from_splits`estract the performances on each distance of the `graubner_nixdorf_sprints` dataset, which are then used to estimate the alactic and lactic capacities of the fictitious athletes.

```{r}
performance_table <- graubner_nixdorf_perf_from_splits()

# filter male performances

male_perf_table <- performance_table |> 
  filter(sex == "male")

male_perf_table

```

The function `sprint_get_capacities_from_perf` estimates the alactic and lactic capacities of the fictitious athletes based on their performances. We won't run the actual function on `male_perf_table`, since the computation take a few minutes, but the user can try the following function call on `male_perf_table`: `sprint_get_capacities_from_perf(male_perf_table)`. It will return an alactic capacity of 328.4066 J/kg and a lactic capacity of 1458.353 J/kg. These values correspond to the ones reported in Briand et al. (2025).

Similarly, the performance table can be filtered to retain only the data from the female athlete. 

```{r}
female_perf_table <- performance_table |> 
  filter(sex == "female")

female_perf_table
```
We can the function to get capacities from sprint performances on `female_perf_data`. Again we won't run the whole function in a dedicated chunk, as it takes several minutes. to run. To reproduce the values obtained in Briand et al. (2025), the user can run the following script:  `sprint_get_capacities_from_perf(female_perf_table, maximal_aerobic_power = 21,  init_alactic = 250, init_lactic = 1300, lower_bounds = c(200, 1100), upper_bounds = c(295, 1400))`. The function provides an alactic capacity of 215.5259 J/kg, and a lactic capacity of 1303.516 J/kg. These values are vey close to the alactic and lactic capacities reported in Briand et al. (2025), which were of 218 J/kg and 1295 J/kg respectively for the alactic and lactic capacities. As before, these differences are due to small modification in the integration methods and time step for the sake of computational efficiency within the package, compared to the initial approach used when performing the package initial computations. For the same reasons, similar small differences between the vignette output and the values reported in the package were already observed in the package vignette `sprint_bioenergetic_model.Rmd`.

As in Briand et al. (2025), we can generate a table that summarizes the fictitious athletes' simulation, including the distance travelled, the error with respect to actual travelled distance, as well as the total energy expenditure and the energy expenditure from the three energy pathways. A table similar to the one presented in Briand et al. (2025) can be generated using the following code. The function default inputs corresponds to the fictitious male and female alactic and lactic capacities, as well as maximal aerobic power, as reported in Briand et al. article. The parameters can be adjusted and differences between the values generated in the table and the ones reported in Briand et al. (2025) are due to rounding of the capacities. 

```{r}
sprint_simulation_metrics_briand_article()
```


## Simulate Athlete Performances From Capacities

The last step step consists in estimating the male and female sprint time over different distances given their alactic and lactic capacities, as well as their maximal aerobic power. The function `sprint_time_perf_simulation` can be used to simulate the sprint time performances of the fictitious athletes on tagerted distances. We can apply the function on the previously estimated alactic and lactic capacities for the fictioous athlete of 328 J/kg and 1460 J/kg, respectiverly, and a maximal aerobic power of 24.5 W/kg, which is the value used in Briand et al. (2025). The function returns a tibble with the target distance, the running time and the actual time, which includes the reaction time.

```{r}
target_distances <- c(60, 100, 200, 300,400)
alactic_capacity <- 328
lactic_capacity <- 1460

# run the simulation

sprint_time_perf_simulation(
  target_distances = target_distances,
  alactic_capacity = alactic_capacity,
  lactic_capacity = lactic_capacity
)


```

The function `sprint_time_perf_simulation` formats the output into a table similar to the one presented in Briand et al. (2025) original artical. We can generate the table on both male and female estimated capacities.

```{r}
# male table
sprint_time_perf_simulation_briand_table(
  target_distances = target_distances,
  alactic_capacity = alactic_capacity,
  lactic_capacity = lactic_capacity
)

# female capacities

female_alactic_capacity <- 218
female_lactic_capacity <- 1295


# generate the table

sprint_time_perf_simulation_briand_table(
  target_distances = target_distances,
  alactic_capacity = female_alactic_capacity,
  lactic_capacity = female_lactic_capacity,
  maximal_aerobic_power = 21 
)

```


