---
title: "Sprint Energetics Exploration"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Sprint Energetics Exploration}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(runrgetics)
library(dplyr)
```

# Sprint Energetics Exploration

The goal of this vignette is to explore simple functions that explore sprint metabolic power to gradually develop a sprint bioenergetic model.

`runrgetics` provides functions to estimate instantaneous metabolic power from sprint data, such as the `graubner_nixdorf_sprints`dataset, which contains time splits and average velocities over different distance intervals from the 100 m, 200 m, and 400 m sprint events at the 2009 World Athletics Championships in Berlin.

A visual representation of the sprint motion data reveals metabolic peak powers attained within the first second of the sprint. These peaks are presumed to result from anaerobic alactic energy sources. The following chunk shows the metabolic power contribution as a function of time for the men's 100 m sprint event. 

```{r}
men_100 <- graubner_nixdorf_sprints  |>
  dplyr::filter(event == "Men's 100 m")

plot_sprint_metabolic_power(
  mean_velocity_splits = men_100$velocity,
  time_splits = men_100$splits,
  distance = men_100$distance,
  reaction_time = men_100$reaction_time[1],
  maximal_velocity = men_100$maximal_velocity[1]
)


```

For the men's 100 m, the metabolic power peak reaches nearly 160 W/kg

```{r}

# extract sprint motion data for the 100 m
sprint_data_100 <- sprint_motion_model_data(
  mean_velocity_splits = men_100$velocity,
  time_splits = men_100$splits,
  distance = men_100$distance,
  reaction_time = men_100$reaction_time[1],
  maximal_velocity = men_100$maximal_velocity[1]
)

# calculate maximal power  
maximal_power_100 <- sprint_maximum_metabolic_power(sprint_data_100)
maximal_power_100
```

To model the energy contribution of the anaerobic alactic, lactic and aerobic metabolism, we can start with some exploration to define the approximate contribution of the energy system over a sprint. 

## Assumptions of the Maximal Anaerobic Alactic and Anaerobic Lactic Capacity

To drive this exploration, estimations have to be made on the maximal anaerobic lactic and maximal anerobic alactic capacities. For world class male and female elite sprinters, we can estimate the maximal anaerobic alactic capacity to range between 310 and 390 J/kg for male and between 230 and 290 J/kg for female athletes. The maximal anaerobic lactic capacity can be estimated to be in the magnitude of 1250 J/kg for male and 1150 J/kg for female, corresponding to a blood lactatate accumulation of 20 mmol/L. The derivation of these estimates is provided in ful lenght in the articles by Briand et al. (2025) and di Prampero and Osgnach (2018).

## Approximate Energy Distribution of the Sprint Acceleration Phase

The use of anaerobic alactic energy is mostly present in the sprint acceleration. It is therefore worth analysing the acceleration and deceleration phase of the sprint separately. 

The following chunk of code shows how functions from `runergetics`can be used to separate the sprint motion data (instantaneous distance, velocity, acceleration and power) into two distinct data frames, one for the acceleration and one for the deceleration.

```{r}
# extract acceleration data

acceleration_data <- sprint_motion_acceleration_data(sprint_data_100)
head(acceleration_data)

# extract deceleration data

deceleration_data <- sprint_motion_deceleration_data(sprint_data_100)
head(deceleration_data)

```

### add plot of acceleration and deceleration phases...


We can now make the assumption that all the energy expanded during the acceleration comes form anaerobic alactic sources. We can estimate this quantity by taking the intergal of the acceleration metabolic power using the `runergetics` function `energy_total`.

```{r}
# Estimate the total energy expanded during the acceleration phase

energy_total(acceleration_data)

```

The total energy in the acceleration (~590 J/kg) is greater than the range of estimated maximal alactic capacity (310-390 J/kg) for World class male sprinters. This suggests that the anaerobic alactic energy system is not the only source of energy during the acceleration phase.

### Adding Aerobic Contribution

While the aerobic contribution is small during the sprint acceleration phase, it nonetheless contributes to the total energy expanded. As in Briand et al. (2025), we can assume aerobic power to follow an exponentially rising function.

$$
P_{aer} = (MAP - BMR) \left(1 - e^{-\frac{t}{k_{aer}}}\right)
$$
Where $MAP$ corresponds to maximal aerobic power and can be fixed at 24.5 W/kg for males and 21 W/kg as in Briand et al. (2025). $BMR$ is the basal metabolic rate and can be fixed at 1.2 W/kg. $k_{aer}$ is the time constant of the exponential function and can be fixed at 23 s (cite di prampero etc.). The aerobic power contribution can be estimated using `sprint_approx_aerobic_power` function.


The following chunk of computes the aerobic and anaerobic (difference between aerobic and total) power contribution to the total metabolic power of the acceleration. 

```{r}
# compute the aerobic power contribution

acceleration_data <- acceleration_data |>
  dplyr::mutate(
    power_aerobic = sprint_approx_aerobic_power(time = time), 
    power_anaerobic = power - power_aerobic
  )


# Compute the aerobic and anaerobic power contribution

energy_aerobic <- energy_aerobic(acceleration_data)
energy_anaerobic <- energy_anaerobic(acceleration_data)


```

During the acceleration phase, the aerobic energy expenditure is estimated to be `r round(energy_aerobic, 2)` J/kg, while the anaerobic energy expenditure is estimated to be `r round(energy_anaerobic, 2)` J/kg. The anaerobic energy expenditure is greater than the maximal anaerobic alactic capacity, suggesting that the anaerobic is not exclusively coming from alactic sources, but also from anaerobic lactic sources.

### Considering Approximate Anaerobic Lactic Contribution

Based on previous work modelling energy contributions in sprint running, it can be assumed that anaerobic lactic power follows an exponential rising function in the sprint acceleration phase (see Briand et al. 2025). The maximum ($P_{an,max}$) of this rising function is attained at the end of the acceleration, around 5 s. It therefore has a very short time constant ($k_{an}$) of 2 s. 

$$
P_{an,accel} = P_{an,max} \left(1 - e^{-\frac{t}{k_{an}}}\right)
$$
The anaerobic lactic contribution can be estimated using the `sprint_acceleration_approx_lactic_power` function.

The alactic contribution can be computed as the difference between the lactic contribution and the anaerobic contribution.

```{r}
#estimate the maximal lactic power as the anaerobic power at the end of the acceleration

maximal_lactic_power <- acceleration_data$power_anaerobic[nrow(acceleration_data)]
# compute the aerobic power contribution

acceleration_data <- acceleration_data |>
  dplyr::mutate(
    power_lactic = sprint_acceleration_approx_lactic_power(time = time, maximal_lactic_power = maximal_lactic_power), 
    power_alactic = power_anaerobic - power_lactic,
    
  )


# Compute the aerobic and anaerobic power contribution

energy_lactic <- energy_lactic(acceleration_data)
energy_alactic <- energy_alactic(acceleration_data)


```

Folowing the previous calculations, during the acceleration phase, the anaerobic lactic energy expenditure is estimated to be `r round(energy_lactic, 2)` J/kg, while the anaerobic alactic energy expenditure is estimated to be `r round(energy_alactic, 2)` J/kg. The anaerobic alactic expenditure is within the estimated maximal alactic capacity, suggesting that the proposed energy distribution for the 3 metabolisms is realistic.

## Approximate Distribution of the Deceleration Phase

Following the acceleration, the aerobic power contribution is expected to continue increasing following the same exponentially rising function. We can assume that the anaerobic alactic power contribution decreases and reach a null contribution at the end of the race. The anaerobic lactic contribution presents the remaining power of the total metabolic power. 

Following these assumptions, the function `sprint_approx_distribution`calculates both the acceleration and deceleration phases approximate distributions to provide a global approximate distribution of the energy systems to total power over the whole sprint.

```{r}
# compute the approximate distribution of the energy systems

sprint_approx_power_distribution <- sprint_approx_power_distributions(sprint_data_100)
head(sprint_approx_power_distribution)

```

We can compute the approximate energy contribution from the three metabolic pathways. 

```{r}
energy_lactic <- energy_lactic(sprint_approx_power_distribution)
energy_alactic <- energy_alactic(sprint_approx_power_distribution)
energy_aerobic <- energy_aerobic(sprint_approx_power_distribution)
energy_total <- energy_total(sprint_approx_power_distribution)
```

From the approximate distribution, the total energy expenditure is estimated to be `r round(energy_total, 2)` J/kg. The anaerobic alactic energy expenditure is estimated to be `r round(energy_alactic, 2)` J/kg, while the anaerobic lactic energy expenditure is estimated to be `r round(energy_lactic, 2)` J/kg. The aerobic energy expenditure is estimated to be `r round(energy_aerobic, 2)` J/kg.

plot approximate power distribution ......

### Relative Energy Contributions

Calcluate relative energy contributions in percentages.....


## Alactic Power Model

Functions can be applied to model the approximate alactic power of the course of the sprint. The following log-normal function can be used to model alactic power.: 

$$
P_{al}(t) = P_{al,max}(T) e^{\frac{-(ln(t) - \mu_{al})^2}{2\sigma_{al}^2}}
$$
Where $P_{al,max}(T)$ is the maximal alactic power for a race of duration $T$, $\mu_{al}$ is a constant seting the peak of the log-normal distribution and $\sigma_{al}$ is a parameter setting the distribution's decay. The function `fit_approx_alactic_power_model` can be used to determine the values of parameters $P_{al,max}(T)$, $\mu_{al}$ and $\sigma_{al}$ to the approximate distribution. Following the determination of $\mu_{al}$ and $\sigma_{al}$, the function `sprint_approx_alactic_power_model` can be used to calculate the alactic power distribution. 


```{r}
# Fiting the alactic power model to the approximate distributions

alactic_parameters <- fit_approx_alactic_power_model(sprint_approx_power_distribution
)

p_al_max <- alactic_parameters$pal_max
mu_al <- alactic_parameters$mu
sigma_al <- alactic_parameters$sigma

```

After fitting the alactic power model to the approximate alactic power distribution, a the maximal alactic power over the sprint is estimated to be `r round(p_al_max, 2)` W/kg. The peak of the log-normal distribution is estimated to be `r round(mu_al, 2)` and the decay of the distribution is estimated to be `r round(sigma_al, 2)`. 

The alactic power can be calculated using the alacic power model.

```{r}
# compute the alactic power from the model

sprint_approx_power_distribution <- sprint_approx_power_distribution |> 
  dplyr::mutate(
    power_alactic_model = sprint_approx_alactic_power_model(
      time = time,
      maximal_alactic_power = p_al_max,
      mu = mu_al,
      sigma = sigma_al
    )
  )

```

The alactic power computed using the model can now be compared to the approximate distribution. 

```{r}

```

 plot approximate distribution compared to power calculated using the model .....

## Lactic Power Model

To model the approximate lactic power output over the course of the sprint, a bi-exponential function can be used. The following function can be used to fit the parameters of the bi-exponential function to the approximate distribution.

$$
P_{la}(t) = P_{la,max}(T) \cdot k_{norm}\cdot \left(1 - e^{-\frac{t}{k_1}} \right) \cdot e^{-\frac{t}{k_2}}
$$
Where $P_{la,max}(T)$ is the maximal lactic power for a sprint race of duration $T$, $k_{norm}$ is a normalizing constant, $k_1$ is the time constant of the first rising exponential and $k_2$ is the time constant of the second decaying exponential. The function `fit_approx_lactic_power_model` can be used to determine the values of parameters $P_{la,max}(T)$, $k_{norm}$, $k_1$ and $k_2$ to the approximate distribution. Following the determination of $k_{norm}$, $k_1$ and $k_2$, the function `sprint_approx_lactic_power_model` can be used to calculate the lactic power distribution. 

```{r}
# Fiting the alactic power model to the approximate distributions

lactic_parameters <- fit_approx_lactic_power_model(sprint_approx_power_distribution
)

p_la_max <- lactic_parameters$p_la_max
k1 <- lactic_parameters$k1
k2 <- lactic_parameters$k2
```

After fitting the lactic power model to the approximate lactic power distribution, a the maximal lactic power over the sprint is estimated to be `r round(p_la_max, 2)` W/kg. The time-constant of the rising exponential is estimated to be `r round(k1, 2)` and the time constant of the second decaying exponential is estimated to be `r round(k2, 2)`. 

The lactic power can be calculated using the lacic power model.

```{r}
# compute the alactic power from the model

sprint_approx_power_distribution <- sprint_approx_power_distribution |> 
  dplyr::mutate(
    power_lactic_model = sprint_approx_lactic_power_model(
      time = time,
      maximal_lactic_power = p_la_max,
      k1 = k1,
      k2 = k2
    )
  )

```

The lactic power computed using the model can now be compared to the approximate distribution. 

```{r}

```

 plot approximate distribution compared to power calculated using the model .....
